Open archaeology, open source?

### 2023-06-14 analysis/_04-what_is_open_archaeology.qmd
Exclude old files from quarto renders
Add headings
Revise fig-cats-plats

## Breakdown of tags and categories

@@

| Category                              | Kind      | Scope                                                                                                                                                                                                        |
|:-----------|:-----------|:------------------------------------------------|
| Packages and libraries                | Software  | Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.                                                                                   |
| Standalone software                   | Software  | Software that may be operated without needing to first access an underlying platform.                                                                                                                        |
| Scripts                               | Software  | Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation. |
| Specifications, protocols and schemas | Documents | A formal data structure or framework intended to be used as a model.                                                                                                                                         |
| Products                              | Documents | Stable outcomes of creative work.                                                                                                                                                                            |
| Guides                                | Documents | An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.                                                                                          |
| Lists and datasets                    | Documents | A series of consistently organized observations assembled with purpose.                                                                                                                                      |

@@

## Breakdown of platforms and languages

@@

Most projects included in open-archaeo are designed to be used as a package or library atop am existing "platform" -- for example a package for a programming language like R, or a plugin for an application like QGIS. The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes. Others create standalone software that can be run independently of such platforms, for example desktop or web apps. A significant amount of projects also comprise of datasets and non-packaged code snippets that have been made available for general use.

```{r plot-categories}
oarch <- mutate(oarch,
                category = recode(category,
                                  "Specifications, protocols and schemas" = "Specifications etc.",
                                  .default = category))
oarch %>%
  drop_na(category) %>%
  group_by(category) %>%
  summarise(n = n()) %>%
  ggplot(aes(fct_reorder(category, n), n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Most popular categories",
    subtitle = "By number of projects",
    caption = "Data: open-archaeo.info",
    x = NULL, y = NULL
  ) +
  theme_oarch()
```

R is overwhelmingly dominant as a platform, closely followed by Python. Plugins for QGIS and mobile apps are also relatively popular. Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates. In the figure below, the darker grey lines are used to break up the bars by primary author, showing that many of the smaller platforms---for example Excel templates---are targeted by only one or two developers. The larger platforms tend to be more diverse.

```{r plot-platforms}
oarch %>%
  drop_na(platform) %>%
  group_by(platform, author1_name) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(total = sum(n)) %>%
  ggplot(aes(fct_reorder(platform, total), n, group = author1_name)) +
  geom_col(colour = "#333333") +
  coord_flip() +
  labs(
    title = "Most popular platforms",
    subtitle = "By number of projects, split by primary author",
    caption = "Data: open-archaeo.info",
    x = NULL, y = NULL
  ) +
  theme_oarch()
```

>>>>

# Open archaeology <!-- Results I (description of what's there) -->

Our analysis documented an the widespread presence of open archaeology projects on GitHub and other online platforms.

@@

| Category                              | Kind      | Scope                                                                                                                                                                                                        |
|:-----------|:-----------|:------------------------------------------------|
| Packages and libraries                | Software  | Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.                                                                                   |
| Standalone software                   | Software  | Software that may be operated without needing to first access an underlying platform.                                                                                                                        |
| Scripts                               | Software  | Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation. |
| Specifications, protocols and schemas | Documents | A formal data structure or framework intended to be used as a model.                                                                                                                                         |
| Products                              | Documents | Stable outcomes of creative work.                                                                                                                                                                            |
| Guides                                | Documents | An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.                                                                                          |
| Lists and datasets                    | Documents | A series of consistently organized observations assembled with purpose.                                                                                                                                      |
:Categorisation scheme used by open-archaeo {#tbl-category-defs}

@@

## Platforms and languages

@@

Most projects included in open-archaeo are designed to be used as a package or library atop am existing "platform" -- for example a package for a programming language like R, or a plugin for an application like QGIS ([@fig-cats-plats]). The designers of this code are basically creating additional functions within the base platform that are useful for archaeological purposes. Others create standalone software that can be run independently of such platforms, for example desktop or web apps. A significant amount of projects also comprise of datasets and non-packaged code snippets that have been made available for general use .

```{r fig-cats-plats}
#| fig-cap: Number of open archaeology projects by category and platform
#| fig-subcap: 
#|   - Categories
#|   - Platforms

oarch |>
  drop_na(category) |>
  group_by(category) |>
  summarise(n = n()) |>
  ggplot(aes(fct_reorder(category, n), n)) +
  geom_col() +
  geom_text(aes(label = category), hjust = 0, nudge_y = 1) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 300)) +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank()
  )

oarch |>
  drop_na(platform) |>
  count(platform) |>
  arrange(n) |>
  ggplot(aes(fct_reorder(platform, n), n)) +
  geom_col(colour = "#333333") +
  geom_text(aes(label = platform), hjust = 0, nudge_y = 1) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 200)) +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank()
  )
```

R is overwhelmingly dominant as a platform, closely followed by Python [@fig-cats-plats]. Plugins for QGIS and mobile apps are also relatively popular. Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates. In the figure below, the darker grey lines are used to break up the bars by primary author, showing that many of the smaller platforms---for example Excel templates---are targeted by only one or two developers. The larger platforms tend to be more diverse.

```{r plot-platforms}
```
